Multi-level modeling with Openflexo/\FML

### 2021-10-05 responseLetter.tex

\bigskip
Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador, Fabien and Antoine

\pagebreak

\Editor{}

\begin{response}{the reviewers have made suggestions for improving the presentation but also raised a number of issues requiring clarification. An overarching theme is the request for a more explicit description of how the multi-level modelling emulation is achieved and what its general advantages and limitations are, including a clarification of what the implications would be of supporting more "levels" than the challenge required.}

We have been more explicit to explain how we use the FML language at the end of the introduction, and in various figures. TODO.

\todo[inline]{We have enhanced the presentation of both, Openflexo/FML and our solution to the challenge so that: 1) the techniques we use to \emph{achieve} multi-level capabilities are more explicit; and, 2) the advantages and disadvantages of using them in order to fulfil the challenge requirements become clearer. Notably, we have added many details in the final paragraphs of the introduction section, improved the Section 2, devoted to the description of OpenFlexo, refactored figures X, Y, Z.}
\end{response}

\begin{response}{We very much welcome a "conventional" response to the challenge but would like to see the respective reviewer questions addressed in a revised version which is explicit as possible about the advantages and the disadvantages of the emulation. You may find the references listed below useful to keep explanations concise by referencing related work (note that reviewer 3’s “Type-Instance pattern” is also known as the “Type Object” pattern). You should also respond to other concerns reviewers raise, either by making changes to the submission or in the form of an authors' response.} 
The related work section has been modified to show that we are not the first conventional response with multi-level emulation, and to improve comparisons and limitations of different approaches.\todo[inline]{I do not understand this answer. We are saying that other approaches are also unconventional?}
\end{response}

\begin{response}{Finally, to achieve consistency with other submissions, could you please change the subtitle to "A Contribution to the Multi-Level Process Challenge"?}

Done.
\end{response}

@@

\pagebreak

\Reviewer{\#1}
\begin{response}{In section 2 Technology the authors describe the foundations of the Openflexo approach. Here it would greatly help the reader to see an example how the approach works for traditional modeling languages. For example, how would you realize a small subset of BPMN with the approach? This could briefly be illustrated so as to give the reader an immediate understanding. Especially, the 'metamodel'-agnostic approach seems very interesting but I would like to see how that is used for traditional modeling approaches.}

The Figure \ref{fig:BPMNSubsetExample} (above) shows how to realize a small subset of BPMN with Openflexo/FML (assuming a metamodel of the full specification of BPMN exists). Concretely, we show  how a VirtualModel (SimplifiedBPMN) can filter an existing model (BPMN) and be instantiated. Note that connection with external sources is not required for our solution to the challenge and is only represented in order to illustrate the capabilities of Openflexo/FML. Therefore, in the article we choose not to give this level of details and focus on the left part of the figure.

\todo[inline]{I've refactored the paragraph above. To be verified.}

### 2021-10-06 8-discussion.tex
add examples in section 6 (discussion)

  The Openflexo approach is level-agnostic: ``levels'' have no specific nature
  and there are no numbered levels. In our solution, concepts of a given level
  are grouped into a virtual model. Inheritance and instantiation allow the
  establishment of relationships between concepts from different levels. See figure~\ref{fig:BPMNSubsetExample}
  and \ref{fig:AcmeFullArchitecture}.

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation. For instance the \texttt{Developer} concept, from \texttt{Acme} metamodel, specializes the \texttt{ActorType} of the base metamodel, and the \texttt{Developer} of the \texttt{Acme} model, is an instance of the metamodel concept. \textcolor{red}{Phrase faite a relire sur Developer\dots}

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models.  Note that in our approach, the
  metamodels are built in an ad hoc way together with the models
  (co-construction), therefore the user also validates on an ad hoc basis.

@@

  %\noteJC{pas trop clair pour moi pour le moment -> N/A ? }
  Due to the nature of our approach and its level-agnosticism, \emph{deep characterization} does not
  apply to our solution. Such a mechanism could probably be encoded by specific behaviors, however,
  it would not be a generic mechanism, making it difficult to reuse for another problem.

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked during model edition and enactment. 
Thus, a part of the verification relies on the fact that users add behaviors when they build
the metamodels and the models. \textcolor{red}{An example of constraint\dots}

>>>>

  The Openflexo approach is level-agnostic: ``levels'' have no specific nature
  and there are no numbered levels. In our solution, concepts of a given level
  are grouped into a virtual model. Inheritance and instantiation allow the
  establishment of relationships between concepts from different levels.
%\ref{fig:BPMNSubsetExample}
%  Figure~\ref{fig:AcmeFullArchitecture} shows three levels: the metamodel, the
%  ACME metamodel and the ACME model.
Figure~\ref{fig:MultilevelArchitecture} shows the multilevel architecture of
our solution.
\noteJC{updated}

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation.
  For instance in the figure~\ref{fig:AcmeFullArchitecture}, the
  \texttt{Developer} concept from \texttt{Acme} metamodel specializes the
  \texttt{ActorType} of the base metamodel, and the \texttt{Developer} of the
  \texttt{Acme} model is an instance of the metamodel concept.
%  \textcolor{red}{Phrase faite a relire sur Developer\dots}
\noteJC{updated}

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models. The section~\ref{sec:model} describes
  how the requirements are captured and how behaviors are implemented with
  \FML: for instance, \texttt{isAuthorizedActor} behavior is needed to enforce
  \textbf{P17} requirement, leading to writing \FML code in \texttt{TaskType}
  and \texttt{Task}. Note that in our approach, the metamodels are built in an
  ad hoc way together with the models (co-construction), therefore the user
  also validates on an ad hoc basis.
\noteJC{updated}

@@

  %\noteJC{pas trop clair pour moi pour le moment -> N/A ? }

  Due to the nature of our approach and its level-agnosticism, \emph{deep
  characterization} does not apply to our solution. Such a mechanism could
  probably be encoded by specific concepts and constraints (behaviors).
  However, it would not be a generic mechanism, making it difficult to reuse
  for another problem.  Plus, programming the notion of \emph{depth} would
  probably imply that a virtual model knows its sub-elements, which is not
  desired and considered as bad design.
\noteJC{updated}

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked during model
edition and enactment. Thus, a part of the verification relies on the fact that
users add behaviors when they build the metamodels and the models. For
instance, \textbf{S6} requirement is constrained by an invariant implemented in
\texttt{CodeArtifact}. It ensures that \texttt{artifactType} has the right type
and that it implements the \texttt{languages} relation.
\noteJC{updated}
